Model Checking Succinct and Parametric\\ One-Counter Automata

### 2023-10-13 version-0-VMCAI/bibliography/icalp.tex
section

\documentclass[times,envcountsame]{llncs}
\usepackage{a4wide}
\usepackage{times}
\usepackage{amsmath,amssymb,eufrak,theorem}%mnsymbol}
\usepackage{enumerate}
\usepackage{complexity}
\usepackage{gastex}

@@

%\usepackage[thmmarks,standard,thref]{ntheorem}

\newtheorem{theorem}{{\bf Theorem}}[section]
\newtheorem{corollary}[theorem]{{\bf Corollary}}
\newtheorem{definition}[theorem]{{\bf Definition}}
\newtheorem{example}[theorem]{{\bf Example}}
\newtheorem{lemma}[theorem]{{\bf Lemma}}
\newtheorem{proposition}[theorem]{{\bf Proposition}}
\newtheorem{remark}[theorem]{{\bf Remark}}
\newtheorem{fact}[theorem]{Fact}

@@

%\usepackage[active]{srcltx}

\iffalse
\newcommand{\problemx}[3]{
\par\noindent\underline{\sc#1}\par\nobreak\vskip.2\baselineskip
\begingroup\clubpenalty10000\widowpenalty10000
\setbox0\hbox{\bf INPUT: }\setbox1\hbox{\bf QUESTION: }
\dimen0=\wd0\ifnum\wd1>\dimen0\dimen0=\wd1\fi
\vskip-\parskip\noindent
\hbox to\dimen0{\box0\hfil}\hangindent\dimen0\hangafter1\ignorespaces#2\par
\vskip-\parskip\noindent
\hbox to\dimen0{\box1\hfil}\hangindent\dimen0\hangafter1\ignorespaces#3\par
\endgroup}

\newcommand{\problemy}[3]{
\par\noindent\underline{\sc#1}\par\nobreak\vskip.2\baselineskip
\begingroup\clubpenalty10000\widowpenalty10000
\setbox0\hbox{\bf INPUT: }\setbox1\hbox{\bf OUTPUT: }
\dimen0=\wd0\ifnum\wd1>\dimen0\dimen0=\wd1\fi
\vskip-\parskip\noindent
\hbox to\dimen0{\box0\hfil}\hangindent\dimen0\hangafter1\ignorespaces#2\par
\vskip-\parskip\noindent
\hbox to\dimen0{\box1\hfil}\hangindent\dimen0\hangafter1\ignorespaces#3\par
\endgroup}

@@

\def\AC{{\mathrm{AC}}}
\def\AG{{\mathsf{AG}}}
\def\AND{{\mathrm{AND}}}
\def\BIN{{\mathrm{BIN}}}
\def\CRR{{\mathrm{CRR}}}
\def\CTL{{\text{CTL}}}
\def\div{{\mathrm{div}}}
\def\dom{{\mathrm{dom}}}
\def\EF{{\mathsf{EF}}}
\def\AX{{\mathsf{AX}}}
\def\EXPTIME{{\mathsf{EXPTIME}}}
\def\inp{{\mathrm{in}}}
\def\leaf{{\mathrm{leaf}}}
\def\MAJ{{\mathrm{MAJ}}}
\def\NC{{\mathrm{NC}}}
\def\Omc{{\mathbb{O}}}
\def\OR{{\mathrm{OR}}}
\def\out{{\mathrm{out}}}
\def\Pmc{{\mathcal{P}}}
\def\PSPACE{{\mathsf{PSPACE}}}
\def\TC{{\mathrm{TC}}}
\def\U{{\mathsf{U}}}
\def\WU{{\mathsf{WU}}}
\def\X{{\mathsf{X}}}
\def\EX{{\mathsf{EX}}}
\def\E{{\mathsf{E}}}

@@

\title{Model Checking Succinct and Parametric One-Counter Automata}
\author{Stefan G\"oller\inst{1} \and
Christoph Haase\inst{2} \and
Jo\"el Ouaknine\inst{2} \and
James Worrell\inst{2}
}

@@

\section{Introduction}

Counter automata, which comprise a finite-state controller together
with a number of counter variables, are a fundamental and widely-studied
computational model. One of the earliest results about counter
automata, which appeared in a seminal paper of Minsky's five decades
ago, is the fact that two counters suffice to achieve Turing
completeness~\cite{Min61}.

Following Minsky's work, much research has been directed towards
studying restricted classes of counter automata and related
formalisms. Among others, we note the use of restrictions to a single
counter, on the kinds of allowable tests on the counters, on the
underlying topology of the finite controller (such as
flatness~\cite{CJ98-cav,LS08-atva}), and on the types of computations
considered (such as reversal-boundedness~\cite{ID06-tcs}). Counter
automata are also closely related to Petri nets and pushdown automata.

In Minsky's original formulation, counters were represented as integer
variables that could be incremented, decremented, or tested for
equality with zero by the finite-state controller. More recently,
driven by complexity-theoretic considerations on the one hand, and
potential applications on the other, researchers have investigated
additional primitive operations on counters, such as additive updates
encoded in binary~\cite{BHIMV06,LS08-atva} or even
in \emph{parametric} form, i.e., whose precise values depend on
parameters~\cite{BIL06-icalp,IJTW93-icalp}. We refer to such counter automata as
\emph{succinct} and \emph{parametric} respectively, the former being
viewed as a subclass of the latter. Natural applications of such
counter machines include the modeling of resource-bounded processes,
programs with lists, recursive or multi-threaded programs, and XML
query evaluation; see, e.g.,
\cite{ID06-tcs,BHIMV06,CR04}.

In most cases, investigations have centered around the decidability
and complexity of the \emph{reachability} problem, i.e., whether a
given control state can be reached starting from the initial
configuration of the counter automaton. Various instances of the
reachability problem for succinct and parametric counter automata are
examined, for example, in~\cite{DG-jlc09,HKOW10,IJTW93-icalp}.

The aim of the present paper is to study the decidability and
complexity of \emph{model checking} for succinct and parametric
one-counter automata. In view of Minsky's result, we restrict our
attention to \emph{succinct one-counter automata (SOCA)} and
\emph{parametric one-counter automata (POCA)}. On the specification
side, we focus on the three most prominent formalisms in the
literature, namely the temporal logics CTL and LTL, as well as the
modal $\mu$-calculus. For a counter automaton $\mathbb{A}$ and a
specification $\varphi$, we therefore consider the question of
deciding whether $\mathbb{A} \models \varphi$,
in case of POCA for all values of the parameters,
%\footnote{In the case of
%parametric automata, one asks whether the automaton satisfies the
%formula \emph{for all} legal values of the parameters.}
and investigate both the \emph{data} complexity (in which the formula
$\varphi$ is fixed) as well as the \emph{combined} complexity of this
problem. Our main results are summarized in Table \ref{T results}.
\begin{table}
\begin{center}
\begin{tabular}{ >{\centering\arraybackslash}m{3cm}
  >{\centering\arraybackslash}m{2cm} |
  >{\centering\arraybackslash}m{3.3cm} |
> {\centering\arraybackslash}m{3.3cm} |}\\
  &  &{\bf SOCA}&{\bf POCA}\\ \hline
{\raisebox{-2ex}[0ex]{{\bf CTL, $\mu$-calculus}}}&   {\bf data} &   & \\ \cline{2-2}
{\raisebox{-2ex}[-5ex]{{$\phantom{\text{CTL}}$}}}&  {\bf combined}&
{\raisebox{2ex}[0ex]{$\EXPSPACE$-complete}} &
{\raisebox{2ex}[0ex]{ $\Pi^0_1$-complete}}\\ \cline{1-4}
{\raisebox{-2ex}[0ex]{{\bf LTL}}}&  {\bf data} &
\multicolumn{2}{c|} {\raisebox{0ex}[0ex]{$\coNP$-complete}}
\\ \cline{2-4}
{\raisebox{-2ex}[-5ex]{{$\phantom{\text{CTL}}$}}}&  {\bf combined}&
$\PSPACE$-complete &
$\coNEXP$-complete\\ \cline{1-4}
\end{tabular}
\ \\[0.4cm]
\caption{The complexity of CTL, the modal $\mu$-calculus, and LTL on SOCA and
POCA.}{\label{T results}}
\end{center}
\end{table}

One of the motivations for our work was the recent discovery that
reachability is decidable and in fact $\NP$-complete for both SOCA and
POCA~\cite{HKOW10}. We were also influenced by the work of Demri and
Gascon on model checking extensions of LTL over non-succinct,
non-parametric one-counter automata~\cite{DG-jlc09}, as well as the
recent result of G\"oller and Lohrey establishing that model checking
CTL on such counter automata is $\PSPACE$-complete \cite{GoLo10}.

We note some interesting differences between our results and
corresponding questions regarding finite automata. For the latter, the
(combined) model checking problems for CTL, the $\mu$-calculus, and
LTL are respectively known to be $\mathsf{P}$-complete, in
$\mathsf{NP} \cap \coNP$, and $\PSPACE$-complete. Somewhat
surprisingly, for SOCA and POCA, the complexity ordering is reversed
and LTL becomes easier to model check than either CTL or the
$\mu$-calculus.

On a technical level, the most intricate result is the
$\EXPSPACE$-hardness of CTL model checking for SOCA, which requires
several steps. We first show that $\mathsf{EXPSPACE}$ is `exponentially
$\mathsf{LOGSPACE}$-serializable', adapting the known proof that
$\PSPACE$ is $\mathsf{LOGSPACE}$-serializable. Unfortunately, and in
contrast to~\cite{GoLo10}, this does not immediately provide
an $\EXPSPACE$ lower bound. In a subsequent delicate stage of the
proof, we show how to `split' the counter in order simultaneously to
perform $\PSPACE$ computations in the counter and manipulate numbers
of exponential size in a SOCA of polynomial size.

Our paper is organized as follows. In Section \ref{S Prelim} we
introduce general notations.
Section \ref{S CTL} deals with CTL and the modal $\mu$-calculus.
LTL model checking is content of Section \ref{S LTL}.
Finally, in Section \ref{S Concl} we give conclusions.

\section{Preliminaries}{\label{S Prelim}}

@@

\newcommand{\Reach}{\text{Reach}}
\newcommand{\Moves}{\text{Moves}}
\newcommand{\bin}{\text{bin}}

By $\Z$ we denote the {\em integers} and by $\N=\{0,1,2,\ldots\}$ the
denote the {\em naturals}.  For each $i,j\in\Z$ we define
$[i,j]=\{k\in\Z\mid i\leq k\leq j\}$ and $[j]=[1,j]$.  For each
$i,n\in\N$, let $\bit_i(n)$ denote the $i^{\text{th}}$ least
significant bit of the binary representation of $n$. Hence
$n=\sum_{i\in\N}2^i\cdot\bit_i(n)$.  By $\bin_m(n)=\bit_0(n)\cdots
\bit_{m-1}(n)$ we denote the first $m$ least significant bits written
from {\em left to right}. When $m$ is not important we just write
$\bin(n)$.
%For any non-empty subset $M\subseteq\Z^+$, define $\gcd(M)$ to be the largest
%$k\geq 1$ that divides every element of $M$.
%We define $\gcd(\emptyset)=0$.
Let $p_i$ denote the $i^{\text{th}}$ prime number for each $i\geq 1$, i.e. $p_1=2$,
$p_2=3$ and so on. We define $\log(n)=\min\{i\geq 1\mid 2^i> n\}$, in other
words $\log(n)$ denotes the number of bits that are needed to represent $n$ in
binary.
All polynomials $p:\N\rightarrow\N$ that occur in this paper are assumed to satisfy
$p(n)\geq n$ for each $n\in\N$.
For each word $v=a_1\cdots a_n\in\Sigma^n$ over some finite alphabet $\Sigma$
and each $i,j\in[n]$ define $v[i,j]=a_i\cdots a_j$ and
$v(i)=v[i,i]$.

\bigskip

\renewcommand{\l}{\triangleright}
\renewcommand{\r}{\triangleleft}
\newcommand{\move}{\vdash}
\noindent
{\bf Turing machines and complexity theory:} In the following, we
introduce $f(n)$ space-bounded deterministic Turing machines (DTMs) in
a suitable way for proving lower bounds. These contain precisely one
input tape and one working tape.  In our setting, the working alphabet is
 assumed to be $\{0,1,\l,\r\}$,
where $\l$ is the {\em left marker}
and $\r$ is the {\em right marker}.  The working tape of the initial
configuration of such a DTM on an input $w\in\Sigma^n$ is
assumed to be $\l0^{f(n)}\r$, whereas its input tape is $\l w\r$.
Before we define DTMs, let $\Upsilon=\{-1,0,+1\}$ denote a set of
directions.

Formally, a {\em $f(n)$-space bounded deterministic Turing machine (DTM)} is a tuple
$\M=(S,\Sigma,s_0,F,\mu)$, where $S$ is a finite set of {\em states},
$\Sigma$ is a {\em finite input alphabet} with $\l,\r\not\in\Sigma$,
$s_0\in S$ is an {\em initial state}, $F\subseteq S$ is a set of {\em final
states}, and with $\Sigma_\M=\Sigma\uplus\{\l,\r\}$ we have that
$$\mu\quad:\quad S\times\Sigma_\M\times\{0,1,\l,\r\}\ \rightarrow\
S\times\Upsilon^2\times\{0,1,\l,\r\}$$
is the {\em transition function}, where $\mu(s,b_1,b_2)=(s',\delta_1,\delta_2,b)$ means
that $\M$ is currently in state $s$, its input head reads $b_1$, its
working tape reads $b_2$, and $\M$ changes to state $s'$, moves its
input head in direction $\delta_1$,
moves its working head in direction $\delta_2$, and
writes the bit $b$.
%We require
%moreover that whenever $\M$'s input (resp. working) head reads $\l$, then $\M$
%must move its input (resp. working) head to the right, i.e. $+1$, and whenever
%$\M$'s input (resp.
%working) head reads $\r$, %it must hold that
%$\M$ moves its
%input (resp. working) head to the left, i.e. $-1$. Formally,
%whenever $\delta(s,b_1,b_2)=(s',\delta_1,\delta_2,b)$, then
%for each $i\in\{1,2\}$ we have that $b_i=\l$ implies $\delta_i=+1$ and $b_i=\r$
%implies $\delta_i=-1$.
Moreover, we require that $\M$ writes a marker precisely when it reads a marker,
more formally: $b_2=m$ if and only if $b=m$ for each $m\in\{\l,\r\}$.
As expected, a {\em configuration of $\M$} is a tuple $(s,\l w\r,\l w'\r,i,j)$,
where $s\in S$ is the current state,
$w\in\Sigma^n$ is an input,
$w'\in\{0,1\}^{f(n)}$ is the content of the working tape of $\M$,
$i\in[0,n+1]$ is the current input head position, and
$j\in[0,f(n)+1]$ is the current working head position.
%For configurations $C_1$ and $C_2$, we write $C_1\move_\M C_2$ whenever $\M$ can
%move from $C_1$ to $C_2$ on one step.
We define the {\em language $L(\M)$ of $\M$} to consist of all words
$w\in\Sigma^*$ such that
the {\em initial configuration}
$(s_0,\l w\r,\l 0^{f(|w|)}\r,0,0)$ reaches a configuration
$(s_0,\l w\r,\l w'\r,i,j)$ satisfying $s\in F$.

By $\L$, $\PSPACE$, and $\EXPSPACE$ we denote the class of all
problems that can be decided by DTM that
is logarithmically, polynomially, exponentially space bounded,
respectively. Recall that $\Sigma^0_1$ (resp. $\Pi^0_1$) is the class of
all languages that are (resp. whose complements are) recursively enumerable.
\newcommand{\Prop}{\mathcal{P}}
\newcommand{\Op}{\text{Op}}
%\newcommand{\add}{\mathsf{add}}
\renewcommand{\A}{\mathbb{A}}
%\newcommand{\EF}{\mathsf{EF}}
%\newcommand{\CTL}{\mathsh{CTL}}
\bigskip

\noindent
\textbf{Transition systems: }
In the following, we fix a countable set of atomic propositions $\Prop$.
A {\em transition system} is a tuple $T=(S,\{S_\prop\mid \prop\in\Prop\},\rightarrow)$, where
$S$ is a set of {\em states}, $S_\prop\subseteq S$ for each $\prop\in\Prop$ and
$S_\prop$ is
non-empty for finitely many $\prop\in\Prop$, and finally
$\rightarrow \subseteq S\times S$ is a set of {\em transitions}.
We prefer to use the infix notation $s_1\rightarrow s_2$ to abbreviate
$(s_1,s_2)\in\rightarrow$.
%A {\em finite path} of length $n$ is sequence of transitions
%$\pi=s_0\rightarrow s_1\cdots \rightarrow s_n$, where $n\in\N $.
 An {\em infinite path} is an infinite sequence
$\pi=s_0\rightarrow s_1\cdots$.
% We denote
%by $|\pi|$ the length of the path $\pi$, which we define to be $\omega$
%if $\pi$ is infinite.
For each infinite path
$\pi= s_0\rightarrow s_1 \rightarrow \cdots$ and
$i\in\N$, we denote by $\pi^i$ the path $s_i\rightarrow s_{i+1}\cdots$
and by $\pi(i)$ the state $s_i$.
Define the {\em trace of $\pi$} as $\trace(\pi):\N\rightarrow2^\Prop$ where
$\trace(i)=\{\prop\in\Prop\mid s_i\in S_\prop\}$ for each $i\in\N$.
\bigskip

\noindent
\textbf{Succinct and parametric one-counter automata: }
A {\em succinct one-counter automaton (SOCA)} is a tuple
 $\Soca=(Q,\{Q_\prop\mid \prop\in\Prop\},E,\lambda)$, where $Q$ is a finite set
of {\em control states}, $Q_\prop\subseteq Q$ for each $\prop\in\Prop$ and
$Q_\rho$ is non-empty for finitely many $\prop\in\Prop$,
$E\subseteq Q\times Q$ is a finite set of {\em transitions}, and
$\lambda:E\rightarrow \Z\cup\{\zero\}$ labels the edges with decrements,
increments, and zero tests.
A {\em parametric one-counter automaton (POCA)} is a tuple
$\Poca(X)=(Q,\{Q_\prop\mid\prop\in\Prop\},E,\lambda)$, where the first
three components are same as for a SOCA,
where $X$ is a finite set of {\em parameters over the natural numbers},
and where $\lambda:E\rightarrow
(\Z\cup\{\zero\}\cup\{\circ x\mid\circ\in\{+,-\}, x\in X\})$.
For each assignment $\sigma:X\rightarrow\N$  the
{\em induced SOCA} is defined as $\Poca^\sigma=(Q,\{Q_\prop\mid\prop\in\Prop\},E,\lambda')$
where $\lambda'(e)=\circ\sigma(x)$ whenever $\lambda(e)=\circ x$
for some $\circ\in\{+,-\}$ and $\lambda'(e)=\lambda(e)$ otherwise.
If $X=\{x\}$ is a singleton, then we also write $\Poca(x)$ for $\Poca(X)$.
The {\em size} of a POCA is defined as
$|\Poca|=|Q|+|X|+|E|\cdot\max\{\log(|a|)\mid a\in\lambda(E)\cap\Z\}$.
Hence, we represent each appearing
integer in binary.
The size of a SOCA is defined analogously.
A SOCA $\Soca=(Q,\{Q_\prop\mid \prop\in\Prop\},E,\lambda)$ describes a transition system $T(\Soca)=(Q\times\N,\{Q_\prop\times\N\mid
\prop\in\Prop\}, \rightarrow)$, where for each $q_1,q_2\in Q$ and each
$n_1,n_2\in\N$ we have $q_1(n_1)\rightarrow q_2(n_2)$ if and only
if either $\lambda(q_1,q_2)=n_2-n_1$, or
($n_1=n_2=0$ and $\lambda(q_1,q_2)=\zero)$.

\section{CTL Model Checking}{\label{S CTL}}

In Section \ref{S CTL syn} we introduce the syntax and semantics of $\ctl$.
Since the upper bounds for the modal $\mu$-calculus follow from
known results and since the lower bounds follow from CTL, we do not introduce
the modal $\mu$-calculus formally.
In Section \ref{S Upper} we state an $\EXPSPACE$ upper bound for the combined
complexity of $\ctl$ and the modal $\mu$-calculus on SOCA and a $\Pi^0_1$ upper bound
on POCA. In Section \ref{S Serial} we recall basic notions from complexity theory
such as serializability and results on Chinese remainder representation.
In Section \ref{S CTL hardness} we prove the main result of this section, namely
that the data complexity of $\ctl$ on SOCA is $\EXPSPACE$-hard.
Finally, we prove a $\Pi^0_1$ lower bound for the data complexity
of $\ctl$ on POCA in Section \ref{S CTL und}.

\subsection{$\ctl$: Syntax and Semantics}{\label{S CTL syn}}
\noindent
Formulas $\varphi$ of $\ctl$ are given by the following
grammar, where $\prop$ ranges over $\Prop$:

$$
\varphi\quad ::=\quad \prop\ \mid \neg\varphi\ \mid\ \varphi\wedge\varphi\ \mid\
\EX\varphi\ \mid \E (\varphi\U\varphi)\ \mid\
\E(\varphi\WU\varphi)
$$

\noindent
We have the following abbreviations:
$\varphi_1\vee\varphi_2=\neg(\neg\varphi_1\wedge\neg\varphi_2)$,
$\varphi_1\rightarrow\varphi_2=\neg\varphi_1\vee\varphi_2$,
$\true=\prop\vee\neg\prop$ for some atomic proposition $\prop\in\Prop$,
and $\EF\varphi=\E(\true\U\varphi)$.
The {\em size} of a $\ctl$ formula is inductively defined as follows:
$|\prop|=1$ for each $\prop\in\Prop$, $|\neg\varphi|=|\EX\varphi|=|\varphi|+1$,
$|\varphi_1\wedge\varphi_2|=|\E(\varphi_1\U\varphi_2)|=|\E(\varphi_1\WU\varphi_2)|=
|\varphi_1|+|\varphi_2|+1$.
Given a transition system $T=(S,\{S_\prop\mid \prop\in\Prop\},\rightarrow)$, a state
$s\in S$, and some $\CTL$ formula $\varphi$, define $(T,s)\models\varphi$ by
induction on the structure of $\varphi$ as follows:

>>>>

\documentclass[times,envcountsame]{llncs}

\usepackage{times}
\usepackage{amsmath,amssymb,eufrak,theorem}%mnsymbol}
\usepackage{enumerate}
\usepackage{complexity}
\usepackage{gastex}
\usepackage{url}

@@

\usepackage{wrapfig}
%\usepackage{floatfig}

@@

\newcommand{\prim}{\text{prime}}

@@

\def\AC{{\mathsf{AC}}}
\def\AG{{\mathsf{AG}}}
\def\AND{{\mathrm{AND}}}
\def\BIN{{\mathrm{BIN}}}
\def\CRR{{\mathrm{CRR}}}
\def\CTL{{\text{CTL}}}
\def\div{{\mathrm{div}}}
\def\dom{{\mathrm{dom}}}
\def\EF{{\mathsf{EF}}}
\def\AX{{\mathsf{AX}}}
\def\EXPTIME{{\mathsf{EXPTIME}}}
\def\inp{{\mathrm{in}}}
\def\leaf{{\mathrm{leaf}}}
\def\MAJ{{\mathrm{MAJ}}}
\def\NC{{\mathrm{NC}}}
\def\Omc{{\mathbb{O}}}
\def\OR{{\mathrm{OR}}}
\def\out{{\mathrm{out}}}
\def\Pmc{{\mathcal{P}}}
\def\PSPACE{{\mathsf{PSPACE}}}
\def\TC{{\mathrm{TC}}}
\def\U{{\mathsf{U}}}
\def\WU{{\mathsf{WU}}}
\def\X{{\mathsf{X}}}
\def\EX{{\mathsf{EX}}}
\def\E{{\mathsf{E}}}

@@

\title{Model Checking Succinct and Parametric\\ One-Counter Automata}
\author{Stefan G\"oller\inst{1} \and
Christoph Haase\inst{2} \and
Jo\"el Ouaknine\inst{2} \and
James Worrell\inst{2}
}

@@

\section{Introduction}
\input{intro.tex}

For reasons of space, we have had to abbreviate or omit a number of
proofs; full details can however be found in the technical
report~\cite{techrep}.

\section{Preliminaries}

@@

\newcommand{\Reach}{\text{Reach}}
\newcommand{\Moves}{\text{Moves}}
\newcommand{\bin}{\text{bin}}
\renewcommand{\l}{\triangleright}
\renewcommand{\r}{\triangleleft}
\newcommand{\move}{\vdash}

%By $\L$, $\PSPACE$, and $\EXPSPACE$
%we denote the class of all problems that can be decided by some deterministic
%Turing machine that
%logarithmically, polynomially, and exponentially space bounded, respectively.
\newcommand{\Prop}{\mathcal{P}}
\newcommand{\Op}{\text{Op}}
%\newcommand{\add}{\mathsf{add}}
\renewcommand{\A}{\mathbb{A}}

By $\Z$ we denote the {\em integers} and by $\N=\{0,1,2,\ldots\}$ the {\em naturals}.
For each $i,j\in\Z$ we define $[i,j]=\{k\in\Z\mid i\leq k\leq j\}$ and $[j]=[1,j]$.
For each $i,n\in\N$, let $\bit_i(n)$ denote the $i^{\text{th}}$ least significant bit of
the binary representation of $n$. Hence $n=\sum_{i\in\N}2^i\cdot\bit_i(n)$.
By $\bin_m(n)=\bit_0(n)\cdots \bit_{m-1}(n)$ we denote the first $m$ least significant
bits written from {\em left to right}.
%When $m$ is not important we just write $\bin(n)$.
Let $p_i$ denote the $i^{\text{th}}$ prime number for each $i\geq 1$.
We define $\log(n)=\min\{i\ge 1 \mid 2^i> n\}$, i.e.
$\log(n)$ denotes the number of bits that are needed to represent $n$ in
binary.
%All polynomials $p:\N\rightarrow\N$ that occur in this paper, are assumed to satisfy
%$p(n)\geq n$ for each $n\in\N$.
For each word $v=a_1\cdots a_n\in\Sigma^n$ over some finite alphabet $\Sigma$
and each $i,j\in[n]$ define $v[i,j]=a_i\cdots a_j$ and
$v(i)=v[i,i]$.
For the rest of the paper, we fix a countable set of {\em atomic propositions} $\Prop$.
A {\em transition system} is a tuple $T=(S,\{S_\prop\mid \prop\in\Prop\},\rightarrow)$, where
$S$ is a set of {\em states}, $S_\prop\subseteq S$ for each $\prop\in\Prop$ and
$S_\prop$ is
non-empty for finitely many $\prop\in\Prop$, and finally
$\rightarrow \subseteq S\times S$ is a set of {\em transitions}.
We prefer to use the infix notation $s_1\rightarrow s_2$ instead of
$(s_1,s_2)\in\rightarrow$.
 An {\em infinite path} is an infinite sequence
$\pi=s_0\rightarrow s_1\rightarrow \cdots$.
% We denote
%by $|\pi|$ the length of the path $\pi$, which we define to be $\omega$
%if $\pi$ is infinite.
For each infinite path
$\pi= s_0\rightarrow s_1 \rightarrow \cdots$ and each
$i\in\N$, we denote by $\pi^i$ the suffix $s_i\rightarrow
s_{i+1}\rightarrow\cdots$
and by $\pi(i)$ the state $s_i$.
% Comment: Formal definition not needed
%Define the {\em trace of $\pi$} as $\trace(\pi):\N\rightarrow2^\Prop$ where
%$\trace(i)=\{\prop\in\Prop\mid s_i\in S_\prop\}$  for each $i\in\N$.
%\noindent
%\textbf{Succinct and parametric one-counter automata: }
A {\em SOCA} is a tuple
 $\Soca=(Q,\{Q_\prop\mid \prop\in\Prop\},E,\lambda)$, where $Q$ is a finite set
of {\em control states}, $Q_\prop\subseteq Q$ for each $\prop\in\Prop$ and
$Q_\rho$ is non-empty for finitely many $\prop\in\Prop$,
$E\subseteq Q\times Q$ is a finite set of {\em transitions}, and
$\lambda:E\rightarrow \Z\cup\{\zero\}$.
We call a SOCA $\Soca$ with $\lambda:E\rightarrow
\{-1,0,1\}\cup\{\zero\}$ a {\em unary one-counter automaton (OCA)}.
A {\em POCA} is a tuple
$\Poca(X)=(Q,\{Q_\prop\mid\prop\in\Prop\},E,\lambda)$, where the first
three components are same as for a SOCA,
$X$ is a finite set of {\em parameters over the naturals},
and $\lambda:E\rightarrow
(\Z\cup\{\zero\}\cup\{-x, +x\mid x\in X\})$.
For each assignment $\sigma:X\rightarrow\N$  the
induced SOCA is defined as $\Poca^\sigma=(Q,E,\lambda')$
where $\lambda'(e)=\sigma(x)$ if $\lambda(e)=x$,
$\lambda'(e)=-\sigma(x)$ if $\lambda(e)=-x$, and
$\lambda'(e)=\lambda(e)$ otherwise.
%% $\lambda'(e)=\circ\sigma(x)$ whenever $\lambda(e)=\circ x$
%% for some $\circ\in\{+,-\}$, $x\in X$, and $\lambda'(e)=\lambda(e)$ otherwise.
If $X=\{x\}$ we also write $\Poca(x)$ instead of $\Poca(X)$.
The {\em size} of a POCA is defined as
$|\Poca|=|Q|+|X|+|E|\cdot\max\{\log(|a|)\mid a\in\lambda(E)\cap\Z\}$.
Hence, we represent each appearing
integer in binary.
The size of a SOCA is defined analogously.
A SOCA $\Soca=(Q,\{Q_\prop\mid \prop\in\Prop\},E,\lambda)$ describes a transition system $T(\Soca)=(Q\times\N,\{Q_\prop\times\N\mid
\prop\in\Prop\}, \rightarrow)$, where for each $q_1,q_2\in Q$ and each
$n_1,n_2\in\N$ we have $q_1(n_1)\rightarrow q_2(n_2)$ iff
either $\lambda(q_1,q_2)=n_2-n_1$, or both
$n_1=n_2=0$ and $\lambda(q_1,q_2)=\zero$.

\section{$\CTL$ Model Checking}
Formulas $\varphi$ of CTL are given by the following
grammar, where $\prop$ ranges over $\Prop$:
\begin{equation*}
\varphi\quad ::=\quad \prop\ \mid \neg\varphi\ \mid\ \varphi\wedge\varphi\ \mid\
\EX\varphi\ \mid \E (\varphi\U\varphi)\ \mid\
\E(\varphi\WU\varphi)
\end{equation*}
\iffalse

\noindent
We have the following abbreviations:
$\varphi_1\vee\varphi_2=\neg(\neg\varphi_1\wedge\neg\varphi_2)$,
$\varphi_1\rightarrow\varphi_2=\neg\varphi_1\vee\varphi_2$,
$\true=\prop\vee\neg\prop$ for some atomic proposition $\prop\in\Prop$, $\false=\neg\true$,
$\EF\varphi=\E(\true\U\varphi)$, and $\AX\varphi=\neg\EX\neg\varphi$.
%The {\em size} of a CTL formula is inductively defined as follows:
%$|\prop|=1$ for each $\prop\in\Prop$, $|\neg\varphi|=|\EX\varphi|=|\varphi|+1$,
%$|\varphi_1\wedge\varphi_2|=|\E\varphi_1\U\varphi_2|=|\E\varphi_1\WU\varphi_2|=
%|\varphi_1|+|\varphi_2|+1$.
\fi
Given a transition system $T=(S,\{S_\prop\mid \prop\in\Prop\},\rightarrow)$, a state
$s\in S$, and some $\CTL$ formula $\varphi$, define $(T,s)\models\varphi$ by
induction on the structure of $\varphi$ as follows:
\begin{eqnarray*}
  (T,s)\models \prop &  \quad\iff\quad & s\in S_\prop\quad \text{ for each
}\prop\in\Prop\\
  (T,s)\models \varphi_1\wedge\varphi_2 &  \quad\iff\quad &
(T,s)\models\varphi_1\text{ and }
(T,s)\models \varphi_2 \\
  (T,s)\models \neg\varphi &  \quad\iff\quad & (T,s)\not\models\varphi\\
  (T,s)\models \EX\varphi &  \quad\iff\quad & (T,t)\models\varphi
\text{ for some } t\in S \text{ with }s\rightarrow t\\
  (T,s)\models \E(\varphi_1\U\varphi_2) &  \quad\iff\quad &
\exists s_0,\ldots,s_n\in S, n\geq 0: s_0=s, (T,s_n)\models\varphi_2, \\
&& \forall i\in[0,n-1]: (T,s_i)\models\varphi_1\text{ and } s_i\rightarrow
s_{i+1}\\
  (T,s)\models \E(\varphi_1\WU\varphi_2) &  \quad\iff\quad &
(T,s)\models\E(\varphi_1\U\varphi_2)\ \text{ or } \exists s_0,s_1,\ldots\in
S.\forall i\geq 0:\\
&& s=s_0,(T,s_{i})\models\varphi_1\text{ and }
s_i\rightarrow s_{i+1}
\end{eqnarray*}

\noindent
Subsequently we use the standard abbreviations for disjunction, and
implication.  Moreover, we define $\true=\prop\vee\neg\prop$ for some
$\prop\in\Prop$ and $\EF\varphi=\E\ (\true\ \U\varphi)$.  Let us
define the {\em CTL model checking problem} on SOCA and POCA
resp.\\[-0.3cm]

\problemx{CTL model checking on SOCA}
{SOCA $\Soca=(Q,E,\lambda)$, $q\in Q$, $n\in\N$ in binary, a CTL formula
$\varphi$.}
{Does $(T(\Soca),q(n))\models\varphi$?}
\ \\[-0.4cm]

\problemx{CTL model checking on POCA}
{POCA $\Poca(X)=(Q,E,\lambda)$, $q\in Q$, $n\in\N$ in binary,  a CTL formula
$\varphi$.}
{Does $(T(\Poca^\sigma),q(n))\models\varphi$ for every $\sigma:X\rightarrow\N$?}

\subsection{Upper Bounds}{\label{S Upper}}
\noindent
%\textbf{Upper bounds:} First, let us state the upper bounds on model
%checking CTL on SOCA and
% Comment: Moved to preliminaries
%% {\em One-counter automata (OCA)} are SOCA in which
%% the numbers that occur in the transition labels are represented  in unary
%% instead in binary.
%% The following theorem gives an upper bound on model checking the
%% {\em modal $\mu$-calculus} on OCA.
Due to space restrictions we do not formally introduce the modal
$\mu$-calculus and refer to \cite{ArnNiw01} for more details instead.
In \cite{Serr06} Serre showed that the combined complexity of the
modal $\mu$-calculus on OCA is in $\PSPACE$.  Since every SOCA can be
transformed into an OCA of exponential size, and since each CTL
formula can be translated into an alternation-free $\mu$-calculus
formula with a linear blowup, the following proposition is immediate
by adjusting the resulting $\mu$-calculus formula appropriately.
Moreover, this immediately implies the containment in the arithmetic
hierarchy of the combined complexity of the modal $\mu$-calculus and
CTL on POCA.

%a $\Pi^0_1$ upper bound for the combined complexity of
%the modal $\mu$-calculus and CTL on POCA.
\begin{proposition}{\label{P CTL up}}
For the modal $\mu$-calculus and CTL the combined complexity on SOCA is in
$\EXPSPACE$, whereas it is in $\Pi^0_1$ on POCA.
\end{proposition}
